$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-02 paper/paper.tex
Better organize related work

\begin{example}[Variable reference]
  \autoref{lst:channels} types a variable reference under a linear algebra.
  \begin{lstlisting}[label=lst:channels,mathescape,numbers=left,caption=Typing variable reference $\suc \constr{0}$ with type $\channel{\unit}{\li}$ and usage $\li$.]
_ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\suc \constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
_ = $\suc \constr{0}$
  \end{lstlisting}
  The underscore \_ introduces an anonymous declaration immediately followed by its definition.
  We must show that the variable at position $\suc \constr{0}$ is of type $\channel{\unit}{\li}$ and has a usage annotation $\li$ in an environment with a typing context $\constr{[]} \comma \channel{\unit}{\li} \comma \unit$ and a usage context $\constr{[]} \comma \lio \comma \lio$.
  Which \textsc{VarRef} constructors to use is completely determined by the $\suc \; \constr{0}$ in the type.
  The constructor $\suc$ steps under the outermost variable in the context, preserving its usage annotation $\lio$ from input to output.
  The constructor $\constr{0}$ asserts that the next variable is of type $\channel{\unit}{\li}$, and that we can split its input annotation $\lio$, taking away $\li$ and leaving $\lo$ as the leftover.
\end{example}

@@

  Notice that the \emph{only} difference between the definition of the untyped process $\func{P}$ and its typing derivation are the extra type annotations that accompany scope restriction.
  This extra information requires a (limited) choice.
  To type $\func{P}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $\one \comma \one$ capable of transmitting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we only transmit input capabilities over it (line 9).
  The left-hand process (line 3) is typed following its syntax (line 10).
  For the right-hand (lines 4-5) we need the channel that is going to be sent (lines 11-12).
  The type $\unit$ of this channel is entirely determined by the type $\channel{\unit}{\func{\omega}}$ of the channel over which it is going to be transmitted.
  Its usage annotation is $\one \comma \one$: we will transmit $\one \comma \zero$ and save $\zero \comma \one$ for the continuation process.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Conclusion, Related and Future Work}

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we follow Allais \cite{Allais2018a} and apply leftover typing to the \picalc{}.
We give a more general usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work, we too are able to talk about fully consumed resources --- e.g., we can transmit over a channel $\lz$ multiplicities of a fully exhausted channel.
%
In their recent paper (unpublished at the time of writing), Ciccone and Padovani mechanise a dependently-typed linear \picalc{} in Agda \cite{Ciccone}.
They do so by defining an intrinsicly-typed calculus, which allows them to leverage Agda to provide a dependently-typed interpretation of messages.
Contexts contain not only types, but also the messages that inhabit their corresponding interpretation in Agda.
To avoid linearity violations, channel types as interpreted as unit types --- and therefore only present in the context in the object language.
This approach allows them to encode message predicates, branching, and variable-length conversations, as message input is modelled as dependent function abstraction in Agda.
They use the concrete set of multiplicities $0$, $1$, $\omega$, define a partial algebra on contexts and require extrinsic proofs of context split.
%
Orchard et al. introduce Granule \cite{Orchard}, a fully-fledged functional language with graded modal types, linear types, indexed types and polymorphism.
Modalities include exact usages, security levels and intervals; resource algebras are pre-ordered semirings with partial addition.
They provide bidirectional typing rules, and show the type safety of their semantics.

Recent years have seen an increase in the efforts to mechanise resource-aware process algebras.
One of the earliest works is the mechanisation of the linear \picalc{} in Isabelle/HOL by Gay \cite{Gay2001}.
Gay encodes the \picalc{} with linear and shared types using de Bruijn indices, a reduction relation and a type system posterior to the syntax.
However, in his work context splits are extrinsic, and used multiplicities erased from context.
We make context splits intrinsic, preserve the ability to talk about used resources, and adopt a more general usage algebra: our framework integrates three different type systems into one.
%
The work by Goto et al. \cite{Goto2016a} is, to the best of our knowledge, the first formalisation of session types that includes a type safety proof (in Coq). 
The authors extend session types with polymorphism and pattern matching.
They use a locally-nameless encoding for variable references, a syntax prior to types, and an LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reordering of contexts and extrinsic context splits. 
%
Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq  --- in Coq de Bruijn indices are not as popular as in Agda or Idris because dependent pattern matching is far from straightforward.
As a use-case, they use their tool to help automate proofs of subject reduction for a type system with session types.
%
Taking an intrinsic approach to typing, Thiemann formalises in Agda the MicroSession (minimal GV \cite{Gay2010}) calculus with support for recursion and subtyping \cite{Thiemann2019}.
Context splits are still given extrinsically, and exhausted resources are removed from typing contexts altogether.
The semantics are given as an intrinsically typed CEK machine with a global context of session-typed channels.
%
Most recently, Rouvoet et al. provide an intrinsic type system for a \lambdacalc{} with session types \cite{Rouvoet2020}.
They use proof relevant separation logic and a notion of a supply and demand \emph{market} to make context splits transparent to the user.
Their separation logic is based on a partial commutative monoid that need not be deterministic nor cancellative.
Their typing rules preserve the balance between supply and demand, and are extremely elegant.
They distill their typing rules even further by modelling the supply and demand market as a state monad.
%
Based on contextual type theory \cite{Pientkaa, Pientka}, LINCX \cite{Georges2017} extends the linear logical framework LLF \cite{Cervesato1996} by internalising the notion of bindings and contexts.
The result is a meta-theory in which HOAS encodings with both linear and dependent types can be described.
The developer obtains for free an equational theory of substitution and decidable typechecking without having to encode context splits within the object language.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
To conclude, in this paper we provide a well scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1500 lines of code in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.
%
As future work, we intend to prove further properties of our type system, such as that reduction preserves the balancing of channels.
We intend to add support for products, sum types and recursion to both our syntax and our type system.
Orthogonally, making our typing rules bidirectional would allow us to provide a decision procedure for type checking processes in a given set of algebras.
Furthermore, it might also be worth identifying correspondences between our usage algebra and particular state machines.
Finally, we will use our \picalc{} with leftovers as an underlying framework on top of which we can implement session types and other advanced type theories.

>>>>

\begin{example}[Variable reference]
  \autoref{lst:channels} types a variable reference under a linear algebra.
  \begin{lstlisting}[label=lst:channels,mathescape,numbers=left,caption=Typing variable reference $\suc \constr{0}$ with type $\channel{\unit}{\li}$ and usage $\li$.]
_ : $\contains{\constr{[]} \comma \channel{\unit}{\li} \comma \unit} {\constr{[]} \comma \lio \comma \lio} {\suc \constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo \comma \lio}$
_ = $\suc \constr{0}$

_ : $\contains{\constr{[]} \comma \channel{\unit}{\li}} {\constr{[]} \comma \lio} {\constr{0}} {\channel{\unit}{\li}} {\li} {\constr{[]} \comma \lo}$
_ = $\constr{0}$
  \end{lstlisting}
  The underscore \_ introduces an anonymous declaration immediately followed by its definition.
  We must show that the variable at position $\suc \constr{0}$ is of type $\channel{\unit}{\li}$ and has a usage annotation $\li$ in an environment with a typing context $\constr{[]} \comma \channel{\unit}{\li} \comma \unit$ and a usage context $\constr{[]} \comma \lio \comma \lio$.
  The \textsc{VarRef} constructors we must use are completely determined by $\suc \; \constr{0}$ in the type.
  The constructor $\suc$ steps under the outermost variable in the context, preserving its usage annotation $\lio$ from input to output, the result is the obligation on line 4.
  The constructor $\constr{0}$ asserts that the next variable is of type $\channel{\unit}{\li}$, and that the usage annotation $\lio$ can be split such that $\op{\lio}{\li}{\lo}$ --- we use $\func{\cdot-compute^r}$ to fulfill this proof obligation automatically.
\end{example}

@@

  Notice that the \emph{only} differences between the definition of the untyped process $\func{P}$ and its typing derivation are the extra type annotations that accompany scope restriction.
  This extra information demands a (limited) choice.
  To type $\func{P}$, we need a channel over which we can transmit another channel.
  We therefore create a new channel with usage annotations $\one \comma \one$ capable of transmitting data of type $\channel{\unit}{\func{\omega}}$, and declare (with $\li$) that we only transmit input capabilities over it (line 9).
  The left-hand process (line 3) is typed following its syntax (line 10).
  For the right-hand (lines 4-5) we need the channel that is going to be sent (lines 11-12).
  The type $\unit$ of this channel is entirely determined by the type $\channel{\unit}{\func{\omega}}$ of the channel over which it is going to be transmitted.
  Its usage annotation is $\one \comma \one$: we will transmit $\one \comma \zero$ and save $\zero \comma \one$ for the continuation process.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Related Work}

\paragraph{Extrinsic Encodings}

Extrinsic encodings define a syntax (often well-scoped) and a runtime semantics prior to any type system.
This allows one to talk about ill-typed terms, and defers the proof of subject reduction to a later stage.

Allais \cite{Allais2018a} uses leftover typing to mechanise in Agda a bidirectional type system for the linear \lambdacalc{}.
He proves type preservation and provides a decision procedure for type checking and type inference.
In this paper, we follow Allais \cite{Allais2018a} and apply leftover typing to the \picalc{}.
We generalise the usage algebra, leading to linear, graded and shared type systems.
Drawing from quantitative type theory (by McBride and Atkey \cite{McBride2016, Atkey2018}), in our work we too are able to talk about fully consumed resources --- e.g., we can transmit $\lz$ multiplicities of a fully exhausted channel.

Recent years have seen an increase in the efforts to mechanise resource-aware process algebras, but one of the earliest works is the mechanisation of the linear \picalc{} in Isabelle/HOL by Gay \cite{Gay2001}.
Gay encodes the \picalc{} with linear and shared types using de Bruijn indices, a reduction relation and a type system posterior to the syntax.
However, in his work typing rules demand user-provided context splits, and variables with consumed usage annotations are erased from context.
We remove the demand for context splits, preserve the ability to talk about consumed resources, and adopt a more general usage algebra.

Orchard et al. introduce Granule \cite{Orchard}, a fully-fledged functional language with graded modal types, linear types, indexed types and polymorphism.
Modalities include exact usages, security levels and intervals; resource algebras are pre-ordered semirings with partial addition.
They provide bidirectional typing rules, and show the type safety of their semantics.

The work by Goto et al. \cite{Goto2016a} is, to the best of our knowledge, the first formalisation of session types to come along with a mechanised proof of type safety (in Coq).
The authors extend session types with polymorphism and pattern matching.
They use a locally-nameless encoding for variable references, a syntax prior to types, and an LTS semantics that encodes session-typed processes into the \picalc{}.
Their type system uses reordering of contexts and extrinsic context splits.

\paragraph{Intrinsic Encodings}

Intrinsic encodings merge syntax and type system.
As a result, one can only ever talk about well-typed terms, and the reduction relation by construction carries a proof of subject reduction.
Significantly, by merging the syntax and static semantics of the object language one can fully use the expressive power of the host language.

Thiemann formalises in Agda the MicroSession (minimal GV \cite{Gay2010}) calculus with support for recursion and subtyping \cite{Thiemann2019}.
As Gay does \cite{Gay2001}, context splits are given extrinsically, and exhausted resources are removed from typing contexts altogether.
The runtime semantics are given as an intrinsically typed CEK machine with a global context of session-typed channels.

In their recent paper, Ciccone and Padovani mechanise a dependently-typed linear \picalc{} in Agda \cite{Ciccone}.
They do so by leveraging Agda's dependent types to provide a dependently-typed interpretation of messages.
Contexts contain not only types, but also the messages that inhabit their corresponding interpretation in Agda.
To avoid linearity violations, the interpretation of channel types is erased --- they are only present in the context of the typing judgment.
They model message input as a dependent function abstraction in Agda, and as a result are able to encode message predicates, branching, and variable-length conversations.
They use the concrete set of multiplicities $0$, $1$, $\omega$, define a partial algebra on contexts and require user-provided proofs of context split.

In another recent work, Rouvoet et al. provide an intrinsic type system for a \lambdacalc{} with session types \cite{Rouvoet2020}.
They use proof relevant separation logic and a notion of a supply and demand \emph{market} to make context splits transparent to the user.
Their separation logic is based on a partial commutative monoid that need not be deterministic nor cancellative.
Their typing rules preserve the balance between supply and demand, and are extremely elegant.
They distill their typing rules even further by modelling the supply and demand market as a state monad.

\paragraph{Other Work}

Castro et al. \cite{Castro2020} provide tooling for locally-nameless representations of process calculi in Coq  --- in Coq de Bruijn indices are not as popular as in Agda or Idris because dependent pattern matching is far from straightforward.
As a use-case, they use their tool to help automate proofs of subject reduction for a type system with session types.

Based on contextual type theory \cite{Pientkaa, Pientka}, LINCX \cite{Georges2017} extends the linear logical framework LLF \cite{Cervesato1996} by internalising the notion of bindings and contexts.
The result is a meta-theory in which HOAS encodings with both linear and dependent types can be described.
The developer obtains for free an equational theory of substitution and decidable typechecking without having to encode context splits within the object language.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Conclusion and Future Work}

In this paper we provide a well scoped syntax and a semantics for the \picalc{}, extrinsically define a type system on top of the syntax capable of handling linear, graded and shared types under the same unified framework and show that reduction preserves the well-typedness of processes.
We avoid the need for extrinsic context splits by defining a type system based on leftover typing \cite{Allais2018a}, which is defined here for the first time for the \picalc{}.
As a result, theorems like framing, weakening and strengthening can now we stated for the linear \picalc{}.
Our work is fully mechanised in around 1500 lines of code in Agda%
\begin{anonsuppress}
\cite{Zalakain2020Agda}
\end{anonsuppress}
.

As future work, we intend to prove further properties of our type system, such as that reduction preserves the balancing of channels.
We intend to add support for products, sum types and recursion to both our syntax and our type system.
Orthogonally, making our typing rules bidirectional would allow us to provide a decision procedure for type checking processes in a given set of algebras.
Furthermore, it might also be worth identifying correspondences between our usage algebra and particular state machines.
Finally, we will use our \picalc{} with leftovers as an underlying framework on top of which we can implement session types and other advanced type theories.
